---
title: "01_Cleaning"
author: "Rachel Yorlets"
date: "10/6/2022"
output: html_document
---
The goal of this code file is to clean and retain sociodemographic variables from the SurveillanceExtended data set.

# Step 1
Load packages
```{r}
pacman::p_load(haven,       # read_dta()
               lubridate,   # dates
               dplyr,       # recode(), rename()
               data.table,  # child code
               questionr,   # rename.variable() in child code
               tidyr        # pivot_longer()
               )
```
# Step 2
Load data set: AHRI.Surveillance Episodes Extended. Release 2019-07 - 2000
Use this path for OSCAR
~ refers to the home folder
```{r}
s <- read_dta("~/AHRI data/AHRI.Surveillance Episodes Extended. Release 2019-07 - 2000/SurveillanceEpisodesExtended.dta")
```
# Step 3
Check min and max dates in the surveillance data set
```{r}
min(s$StartDate)
max(s$EndDate)
# Surveillance data set extends from 01Jan2000 to 31Dec2018
```
# Step 4
Create a copy of the data set
```{r}
# Rename data sets
data <- s
```
# Step 5
Prepare key variables that are needed in the data set
```{r}
# Calculate elapsed time in each episode
data$elapsed.time <- elapsed.time <- data$StartDate %--% data$EndDate

# Denote factor variables
data$Education <- as_factor(data$Education)
data$Sex <- as_factor(data$Sex)
data$MaritalStatus <- as_factor(data$MaritalStatus)
```
# Step 6
Select variables to retain in data frame
```{r}
# Select variables
d <- data %>%
  select(IndividualId, Episode, Resident, LocationId, elapsed.time, Year, StartDate, EndDate, HIVStatus, EarliestHIVNeg, LatestHIVNeg, EarliestHIVPos, LatestHIVPos, ARTInitiation, EarliestARTInitDate, DoD, Died, Age, Sex, HouseholdId, HHRelation, Education, MaritalStatus, SEIdx, Cause1, Lik1, Cause2, Lik2, Cause3, Lik3, ImmediateCause, UnderlyingCause, AssociatedCause1, AssociatedCause2, LostToFollowUp, Refusal)

length(unique(d$IndividualId))
#220,648

#View(d[, c('IndividualId', 'Episode', 'EarliestHIVPos', 'EarliestARTInitDate')])
```
# Step 7
Clean sex variable
```{r}
#Sex
#Cannot drop 'unknown' since there are two people in this category
#d$Sex <- droplevels(d$Sex)
# > length(unique(d$IndividualId[d$Sex == 'Unknown']))
# [1] 2
```
# Step 8
Clean education variable
```{r}
#Education
length(is.na(unique(d$IndividualId[d$Education])))
length(is.na(d$IndividualId[d$Education]))

table(d$Education)

d$edu <- recode(d$Education,
                #old = new
                'None' = 'Grade 7 or lower',
                'Grade 1' = 'Grade 7 or lower',
                'Grade 2' = 'Grade 7 or lower',
                'Grade 3' = 'Grade 7 or lower',
                'Grade 4' = 'Grade 7 or lower',
                'Grade 5' = 'Grade 7 or lower',
                'Grade 6' = 'Grade 7 or lower',
                'Grade 7' = 'Grade 7 or lower',
                'Grade 8' = 'Grade 8 or higher',
                'Grade 9' = 'Grade 8 or higher',
                'Grade 10' = 'Grade 8 or higher',
                'Grade 11' = 'Grade 8 or higher',
                'Grade 12' = 'Grade 8 or higher',
                'Some tertiary' = 'Grade 8 or higher',
                'Tertiary qualification' = 'Grade 8 or higher')
#to drop the '?' category, which has no observations in it
d$edu <- droplevels(d$edu)
table(d$edu, exclude = NULL)
```
# Step 9
Clean marital status variable
```{r}
#Marital status
d$marital <- recode(d$MaritalStatus,
                    # old=new
                    'Unknown' = '<NA>',
                    'Not married' = 'Not married',
                    'Married' = 'Married or cohabiting',
                    'Widowed' = 'Not married',
                    'Separated/divorced' = 'Not married',
                    'Informal union' = 'Married or cohabiting')
d$marital <- droplevels(d$marital)
table(d$marital, exclude = NULL)
```
# Step 10
To prepare for creating the time-varying number of children, which involves a Cartesian merge, set unnecessary data sets/copies = NULL
```{r}
data <- NULL
```
# Step 11
Calculate time-varying number of children in household for each person in the full surveillance data set
Maxime Inghels 
```{r}
#Convert d to a data.table to also help conserve memory
d <- as.data.table(d)

sepext2019 <- d
d <- NULL

#Rename IndividualId
sepext2019 <- sepext2019 %>%
  rename.variable(old = 'IndividualId', new = 'IIntId')

# select age<15
setDT(sepext2019)
dt <- sepext2019[Age < 15,.(IIntId, HouseholdId, StartDate, EndDate)]

# remove duplicated lines (when an event occur the next day)
dt <- as.data.table(pivot_longer(dt, cols = c("StartDate", "EndDate")))
setorder(dt, HouseholdId, IIntId, value, -name)

dt[, RowNumber := 1:.N]

tmp <- dt[(name == "StartDate" & IIntId == lag(IIntId) & HouseholdId == lag(HouseholdId) & 
             (value == lag(value) | value == lag(value)+1)) |
            (name == "EndDate" & IIntId == lead(IIntId) & HouseholdId == lead(HouseholdId) & 
               (value == lead(value) | value == lead(value)-1)), ]

dt <- dt[!RowNumber %in% tmp$RowNumber,]
dt$RowNumber <- NULL

# count the number of household member <15
setorder(dt, HouseholdId, value, -name)

dt[, tmp := ifelse(name == "StartDate", 1, -1)]
dt <- dt[, sum(tmp), by = .(HouseholdId, value)]

dt[, ChildNumber := cumsum(V1), by = HouseholdId]
dt$V1 <- NULL
dt <- rename.variable(dt, old = "value", new = "StartDate")

# add EndDate and link with max date follow-up
## EndFollowUp
dt <- merge(dt, sepext2019[ ,.(EndFollowUp = max(EndDate),
                              StartFollowUp = min(StartDate)), 
                           by = HouseholdId], by = "HouseholdId", all.x = T)

setorder(dt, HouseholdId, StartDate)
dt[, EndDate := lead(StartDate), by = HouseholdId]

dt <- dt[StartDate != EndFollowUp,]
dt[is.na(EndDate), EndDate := EndFollowUp]

dt[StartDate == lag(EndDate), StartDate := StartDate+1]
dt$EndFollowUp <- NULL

## StartFollowUp, had record for those who start follow-up without children
dt[,StartDateMin := min(StartDate), by = HouseholdId]
id <- dt[StartDateMin != StartFollowUp, HouseholdId]

dt <- rbind(unique(dt[HouseholdId %in% id, .(HouseholdId, StartDate = StartFollowUp, 
                                             EndDate = as.Date(NA), ChildNumber = 0)]),
      dt[,.(HouseholdId,StartDate,EndDate,ChildNumber)])

setorder(dt, HouseholdId, StartDate)

dt[, tmp:= lead(StartDate)-1, by = HouseholdId]
dt[is.na(EndDate), EndDate := tmp]
dt$tmp <- NULL

# add household with no children during all the follow-up
id <- unique(sepext2019$HouseholdId)[!unique(sepext2019$HouseholdId) %in% unique(dt$HouseholdId)]

dt <- rbind(dt, unique(sepext2019[HouseholdId %in% id, .(HouseholdId,
                                                         StartDate = min(StartDate), 
                                                         EndDate = max(EndDate), 
                                                         ChildNumber = 0)]))

# keep only required variables
dt <- dt[, .(HouseholdId, Start = StartDate, End = EndDate, ChildNumber)]

#Select only essential columns from surveillance data set subset
sep <- sepext2019[,.(IIntId, HouseholdId, StartDate, EndDate, Episode)]

#clean to prepare for merge
#sepext2019 <- NULL
gc()

################
#Merge from Maxime
#merge sep with dt + build final dataset
sep <- merge(sep, dt, by = 'HouseholdId', allow.cartesian = TRUE, all.x = TRUE)

sep <- sep[StartDate <= End & EndDate >= Start, ]

sep[StartDate < Start, StartDate:= Start]
sep[EndDate > End, EndDate:= End]

setkey(sep, 'IIntId', 'StartDate')
###################

#Checking results---

sep[Episode >1 & IIntId == lag(IIntId) & !StartDate-1 == lag(EndDate) ,] 
# above some have history gaps between two households
sep[Episode >1 & HouseholdId  == lag(HouseholdId) & IIntId == lag(IIntId) & !StartDate-1 == lag(EndDate) ,] 
# above 391 with gaps within one household, but look like it is the same in the sepext2019 dataset

length(unique(sep$IIntId)) == length(unique(sepext2019$IIntId)) # same ID in sep and the original one
nrow(sep[Episode == 1, .N, by = IIntId]) == length(unique(sep$IIntId)) # All individuals have their first episode
sep[IIntId == lead(IIntId) & !(Episode == lead(Episode) | Episode+1 == lead(Episode)), .N] # all the Episodes follows one after another so missing episode

tmp <- merge(sep[,.(V1 = max(Episode)), by = IIntId], sepext2019[,.(V2 = max(Episode)), by = IIntId], by = "IIntId")
tmp[V1 !=V2,.N]# last episode of individuals match

#this is where I paused at 2:03
#Merge back to surveillance data set (other covariates)
df <- merge(sepext2019[, c('StartDate', 'EndDate'):= NULL],
                sep[ ,.(HouseholdId, IIntId, StartDate, EndDate, Episode, ChildNumber)],
                by = c("IIntId", "Episode"),
                all.x = T)

df <- as.data.frame(df)
length(unique(df$IIntId))
#[1] 220648
length(unique(df$IIntId[is.na(df$ChildNumber)]))
#0 persons have NA for missing number of children

setDT(df)

#Arrange by IndividualId and StartDate
setkey(df, 'IIntId', 'StartDate')
#re-number episode numbers so that there are no more duplicates (which were created from splicing person-time for children count)
df[, Episode := 1:.N, by = IIntId]

df <- as.data.frame(df)

# rename IIntId as IndividualId
# newname = oldname
df <- rename(df, IndividualId = IIntId)

# #Check on why 30000 people have NA for ChildNumber
# df$HouseholdId.x[is.na(df$ChildNumber)]
# 
# df$HouseholdId.y[is.na(df$ChildNumber)]
# df$ChildNumber[is.na(df$HouseholdId.y)]
# 
# df_missingchildren <- subset(df, is.na(ChildNumber))
# #look at IndividualId 72 and 153, for example
# 
# View(df_missingchildren[ , c('IndividualId', 'HouseholdId.x', 'HouseholdId.y', 'ChildNumber', 'Episode', 'StartDate', 'EndDate',  'Education', 'Age', 'SEIdx', 'MaritalStatus', 'Sex')])
# 
# df$ChildNumber <- replace_na(df$ChildNumber, 0)
# #look at IndividualId 72 and 153, for example, which we know from above df_missingchildren had missing ChildNumber and missing HouseholdId.y
# View(df[ , c('IndividualId', 'HouseholdId.x', 'HouseholdId.y', 'ChildNumber', 'Episode', 'StartDate', 'EndDate',  'Education', 'Age', 'SEIdx', 'MaritalStatus', 'Sex')])


# #Clean
# d <- NULL
# dchildren <- NULL
```
# Step 12
Run checks to see if child number code is working
Specifically, select 10 individuals, and for each individual, check that the episodes are correct regarding both the surveillance episode and the data set with the # of children when compared to the original surveillance data set
```{r}
df_sub <- subset(df, IndividualId == 11 | IndividualId == 12 | IndividualId == 13 | IndividualId == 14 | IndividualId == 15 | IndividualId == 16 | IndividualId == 17 | IndividualId == 18 | IndividualId == 19 | IndividualId == 20)

s_sub <- subset(s, IndividualId == 11 | IndividualId == 12 | IndividualId == 13 | IndividualId == 14 | IndividualId == 15 | IndividualId == 16 | IndividualId == 17 | IndividualId == 18 | IndividualId == 19 | IndividualId == 20)

View(df_sub[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'ChildNumber', 'Education', 'Age', 'SEIdx', 'MaritalStatus', 'Sex')])

View(s_sub[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'Education', 'Age', 'SEIdx', 'MaritalStatus', 'Sex')])
# Reviewed these checks in lab notebooks for first 10 episodes in each data set for the first 7 people, and covariates all matched by time periods (episode numbers have changed from original s_sub to edited df_sub), with the exception of the children, which changed
#Adding the children added about 1.1M rows of data, and this is just because it spliced existing episodes because children changed within episodes - so we could be more granular
```
# Step 13
Clear memory space by setting data frames <- NULL if we don't need them anymore
```{r}
s <- NULL
s_sub <- NULL
df_sub <- NULL
```
# Step 14
Save data frame with cleaned sociodemographic variables as a .Rdata file
```{r}
Step01Clean <- df
save(Step01Clean, file = '01_Cleaning.Rdata')
```
# Step 15
Clear memory space by setting final data frame <- NULL since we don't need it anymore
```{r}
df <- NULL
```