---
title: "02_DerivingSample"
author: "Rachel Yorlets"
date: "10/6/2022"
output: html_document
---
The goal of this code file is to 
-Load and merge additional data sets for viral load, CD4 from ARTemis, and CD4
from NHLS/TIER.net/ARTemis
-Create the analytic sample (pre-censoring)
-Creates EarliestARTInitPreUTT and EarliestARTInitPostUTT

# Step 1
Load packages
```{r}
pacman::p_load(haven,      # read_dta()
               lubridate,  # date
               dplyr,      # case_when()
               Hmisc,      # Lag()
               tableone,   # CreateTableOne()
               writexl,    # Save table as .xlsx file
               paletteer,  # color palette
               ggplot2,
               stringr,    # use with ggplot
               scales,     # use with ggplot
               mice        # multiple imputation via chained eqns
               )
```
# Step 2
Load data from first coding file: 01_Cleaning.Rmd
```{r}
load('01_Cleaning.Rdata')
CleanSociodem <- as.data.frame(Step01Clean)

length(unique(CleanSociodem$IndividualId))
# 220648
```
# Step 3a
Load data: AHRI Population Intervention Programme Community Viral Load - 2011
```{r}
vl <- read_dta("~/AHRI data/AHRI Population Intervention Programme Community Viral Load - 2011/CVL 01-01 AHRI.PIP.CommunityViralLoad.2020.dta")
```
# Step 3b
Load data set: CD4 count from ARTemis
```{r}
#cd4ARTemisdata <- read_dta('~/AHRI data/ARTemis/ARTemis_CD4.dta')
# make copy of data
# select all columns except for EarliestARTInitDate, because we will merge it to
# a data set that also has EarliestARTInitDate
#cd4data <- cd4ARTemisdata %>%
#  select(IndividualId, TestDate, CD4Count, CD4Pct)

# Set cd4data object as a data frame
#cd4data <- as.data.frame(cd4data)
```
# Step 3c
Load data set: CD4 count from NHLS/TIER.net/ARTemis (data from Dickman Gareta June 2025)
```{r}
cd4nhlsdata <- read_dta('~/AHRI data/ARTemis_CD4_V2.dta')
cd4nhlsdata <- as.data.frame(cd4nhlsdata)
```
# Surveillance data set
# Step 4
Remove all episodes in which person-time is contributed at age < 15 because only 
persons 15 and older are eligible for HIV testing
```{r}
d1 <- subset(CleanSociodem, Age >= 15)

length(unique(d1$IndividualId))
# 158730 persons in the full surveillance data set with episodes in which they 
# were 15 or older
```
# Step 4b | STROBE
How do we go from 220648 to 158730 persons?
```{r}
d1_strobe <- subset(CleanSociodem, Age < 15)

length(unique(d1_strobe$IndividualId))
# 106,236 - the problem with this is that some individuals contributed both 
# person-time that was under and over 15 
# 220648-158730 = 61918 persons who did not contribute person-time at age 15 or 
# older during this time period
```
# Step 5
Restrict to persons who are not missing EarliestARTInitDate
```{r}
d2 <- subset(d1, (!is.na(EarliestARTInitDate)))

length(unique(d2$IndividualId))
# 16905 persons >= age 15 do not have a missing EarliestARTInitDate
```
# Step 5b | STROBE
How do we go from 158730 to 16905 persons?
```{r}
# Within the age >=15 data set (before subset to remove persons with missing 
# EarliestARTInitDate), how many persons have a missing EarliestARTInitDate?
length(unique(d1$IndividualId[is.na(d1$EarliestARTInitDate)]))
# 141825 persons have a missing EarliestARTInitDate

158730-141825
# 16905 persons >= age 15 do not have a missing EarliestARTInitDate
```
# Step 6
Restrict to persons who initiated ART between 01Jan2011 and 31Dec2018
Note: We selected this time period for FUP because these are the years during 
which outcome data (ie, viral load) are clean.
```{r}
# Within persons with a non-missing EarliestARTInitDate, filter to those with an EarliestARTInitDate between dates 01Jan2011 and 31Dec2018
d3 <- subset(d2, ((EarliestARTInitDate >= '2011-01-01') & (EarliestARTInitDate <= '2018-12-31')))

length(unique(d3$IndividualId))
# 10981 persons >= age 15 not missing EarliestARTInitDate AND initiated ART between 01Jan2011 and 31Dec2018
```
# Step 6b | STROBE
How do we go from 16905 to 10981?
```{r}
# Within persons with a non-missing EarliestARTInitDate (ie, data set d2), how 
# many have an EarliestARTInitDate before 01Jan2011?
length(unique(d2$IndividualId[d2$EarliestARTInitDate < '2011-01-01']))
# 5890 persons have an EarliestARTInitDate that comes before 01Jan2011

# Within persons with a non-missing EarliestARTInitDate (ie, data set d2), how 
# many have an EarliestARTInitDate after 31Dec2018?
length(unique(d2$IndividualId[d2$EarliestARTInitDate > '2018-12-31']))
# 34 persons have an EarliestARTInitDate that comes after 31Dec2018

16905-5890-34
# 10981
16905-10981
# 5924 persons have an EarliestARTInitDate that comes before or after the 
# interval 11Jan2011 and 31Dec2018
```
# Step 7
Merge the viral load data set to the surveillance set by 'IndividualId'
```{r}
# Rename column in viral load data set from IIntID to IndividualId 
# (its name in Surveillance)
vl <- vl %>%
  rename(IndividualId = IIntID)

# Merge long-format surveillance with long-format viral load data, using the 
# surveillance data set as the left one (keep all of its rows and add columns 
# from viral load data set onto them)
df1 <- left_join(x = d3, y = vl, by = 'IndividualId') 
```
# Step 8
Manipulate merged data set (viral load data and surveillance data set) to retain 
the variables from the viral load data set with the appropriate rows of the 
surveillance data set
```{r}
df2 <- df1 %>%
  # do not need to group by IndividualID because the viral load variables are all already in rows with the correct IndividualId
  # create an indicator variable 'keep' to indicate if the variables from the viral load data set that are included in a given row of the surveillance data set should be kept or not
  mutate(keep = ifelse(((StartDate <= VisitDate) & (VisitDate <= EndDate)), 1, 0),
         # if they should be kept, the 
         VisitDate = ifelse(keep == 1, as.Date(VisitDate), NA), 
         BMFStatusObservation = ifelse(keep == 1, BMFStatusObservation, NA),
         HIVSampleId = ifelse(keep == 1, HIVSampleId, NA),
         VLDetectableLimit = ifelse(keep == 1, VLDetectableLimit, NA),
         VLResult = ifelse(keep == 1, VLResult, NA),
         VLResultCode = ifelse(keep == 1, VLResultCode, NA))

class(df2$VisitDate) <- 'Date'

# nrow(df2)
# 780792

# View(df2[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'BMFStatusObservation', 'HIVSampleId', 'VLDetectableLimit', 'VLResult', 'VLResultCode')])
```
# Step 9
Remove rows in which IndividualId, Episode, and VisitDate are duplicated
```{r}
df3 <- distinct(df2, IndividualId, Episode, VisitDate, .keep_all = TRUE) 

nrow(df3)
# 442258

# View(df3[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'BMFStatusObservation', 'HIVSampleId', 'VLDetectableLimit', 'VLResult', 'VLResultCode')])
# Note that some episodes are repeated because one row has the viral load variables while the other row is identical EXCEPT for the viral load variables
```
# Step 10
Within episodes (we will group by IndividualId and Episode), we will arrange by 
VisitDate and keep only the first row of each episode group 
(ie, the one with the non-missing VisitDate).
This addresses what we saw in the prior step in which some episodes for a given 
IndividualId are repeated because one row has the viral load variables while the 
other row is identical EXCEPT for the viral load variables.
```{r}
df4 <- df3 %>%
  group_by(IndividualId, Episode) %>%
  arrange(VisitDate) %>%
  filter(row_number() == 1) %>%
  arrange(IndividualId, Episode)

length(unique(df4$IndividualId))
# 10981

# View(df4[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'VisitDate', 'keep', 'BMFStatusObservation', 'HIVSampleId', 'VLDetectableLimit', 'VLResult', 'VLResultCode')])
```
# Step 11
Assign any viral load data from 2016 as missing because the lab values were 
unusable/off according to Frank Tanser
```{r}
df5 <- df4 %>%
  # No need to group_by(IndividualId) or arrange rows because the viral load variables are all already in rows w/the correct IndividualId
  # Create an indicator variable to flag if a VisitDate was captured during 2016, the year in which we will set all vl measurements to NA
  mutate(visit_yr2016 = ifelse(year(VisitDate) == 2016, 1, 0),
         # If a viral load measurement occurred during 2016, we set all associated viral load variables equal to NA
         VisitDate = ifelse(visit_yr2016 == 1, NA, as.Date(VisitDate)),
         BMFStatusObservation = ifelse(visit_yr2016 == 1, NA, BMFStatusObservation),
         HIVSampleId = ifelse(visit_yr2016 == 1, NA, HIVSampleId),
         VLDetectableLimit = ifelse(visit_yr2016 == 1, NA, VLDetectableLimit),
         VLResult = ifelse(visit_yr2016 == 1, NA, VLResult),
         VLResultCode = ifelse(visit_yr2016 == 1, NA, VLResultCode))

class(df5$VisitDate) <- 'Date'

# print(unique(df5$IndividualId[df5$visit_yr2016 == 1]))

# for example, person 33 episode 56 had a VisitDate during 2016, which we can see in this data set
# When we look at person 33 episode 56, we see that visit_yr2016 == 1, but all vl variables are NA, which is what we assigned them to be
length(unique(df5$IndividualId))
# 10981

# View(df5[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'VisitDate', 'visit_yr2016', 'BMFStatusObservation', 'HIVSampleId', 'VLDetectableLimit', 'VLResult', 'VLResultCode')])
```
# Step 12
Denote VLResultCode as a factor variable, and note what codes indicate
```{r}
df5$VLResultCode <- as_factor(df5$VLResultCode)

# df5$VLResultCode_num <- recode(df5$VLResultCode,
#                            #old = new
#                            'Valid result' = 1,
#                            'Below detectable limit' = 2,
#                            'Invalid result' = 3,
#                            'Insufficient for testing' = 4,
#                            'Specimen not found' = 5,
#                            'Excluded from testing due to insufficient funding or field error' = 6,
#                            'Other' = 7)
# df4$VLResultCode_num <- replace_na(df4$VLResultCode_num, 0)
# View(df5[, c('IndividualId', 'Episode', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'VisitDate', 'VLResultCode')])
```
# Step 13
Identify persons who achieved initial viral suppression before EarliestARTInitDate, so that they can be removed from analytic sample
```{r}
df5 <- df5 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # If VisitDate is not missing and occurs before EarliestARTInitDate and the viral load measurement was coded as below detectable limit
  # We use <= EarliestARTInitDate because we do not allow people to become suppressed on the same date they start ART (just not before)
  # We do not need to rank the occurrence of VLResultCode==2 because any occurrence disqualifies them from inclusion
  mutate(suppression_pre_ART = ifelse(((!is.na(VisitDate)) & (VisitDate <= EarliestARTInitDate) & (VLResultCode == 2)), 1, 0)) %>%
  mutate(total_sup_per_person = length(which(suppression_pre_ART == 1)))

length(unique(df5$IndividualId))
# 10981 

length(unique(df5$IndividualId[is.na(df5$suppression_pre_ART )]))
# 0 

# View(df5[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'VisitDate', 'EarliestARTInitDate', 'VLResultCode', 'suppression_pre_ART', 'total_sup_per_person')])
```
# Step 14b | STROBE
How do we go from 10981 to 10739?
```{r}
########
# Before we dropped persons who had the event before EarliestARTInitDate
########

# How many had had a missing value for the event?
length(unique(df5$IndividualId[is.na(df5$suppression_pre_ART)]))
#0

# How many had the event before EarliestARTInitDate?
length(unique(df5$IndividualId[df5$suppression_pre_ART == 1]))
#242 had the event before EarliestARTInitDate, before we subset to people who contributed any follow-up time (ie, they have an origin) or subset to people who contributed person-time for the year prior to EarliestARTInitDate

# Consider the break-down of how many instances of suppression pre-ART occurred among persons who had the event before EarliestARTInitDate
length(unique(df5$IndividualId[df5$total_sup_per_person == 1]))
# 202

length(unique(df5$IndividualId[df5$total_sup_per_person == 2]))
# 36

length(unique(df5$IndividualId[df5$total_sup_per_person == 3]))
# 4

length(unique(df5$IndividualId[df5$total_sup_per_person > 3]))
# 0
# 202+36+4=242
```
# Step 14
Retain only the persons who did not have the event before their EarliestARTInitDate
```{r}
df6 <- df5 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # retain only the persons who never had the event before EarliestARTInitDate (that we know of)
  filter(length(which(suppression_pre_ART == 1)) == 0)
  # remember we have not yet removed the rows that come before EarliestARTInitDate

length(unique(df6$IndividualId))
# 10739 did not have the event before their EarliestARTInitDate

nrow(df6)
# 418642

# We can confirm it this way
df6_confirm <- df5 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # retain only the persons who never had the event before EarliestARTInitDate (that we know of)
  filter(total_sup_per_person == 0)
  # remember we have not yet removed the rows that come before EarliestARTInitDate

length(unique(df6_confirm$IndividualId))
# 10739 did not have the event before their EarliestARTInitDate

# View(df6[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'VisitDate', 'EarliestARTInitDate', 'VLResultCode', 'suppression_pre_ART')])
```
##########################
# Merge CD4 data from NHLS, TIER.net, ARTemis from Dickman Gareta June 2025
#########################
# Step 
First, rename the variables in this data set because it has the same column 
names as the ARTemis-only CD4 data that we just merged above
```{r}
names(cd4nhlsdata) <- c('IndividualId', 'TestDateV2', 'CD4CountV2', 'CD4PctV2', 'DataSourceV2')
```
Merge the CD4 count data set to the merged surveillance + viral load set by 'IndividualId'
```{r}
# Merge long-format CD4 data with long-format surveillance + viral load data, using the surveillance + viral load data set as the left one (keep all of its rows and add columns from cd4 count data set onto them)
survvlcd4V2_1 <- left_join(x = df6, y = cd4nhlsdata, by = 'IndividualId')

# many to many merge
```
# Step 
Manipulate merged data set (viral load data + surveillance data set now merged with cd4 data) to retain the variables from the cd4 data set with the appropriate rows of the surveillance + viral data set
```{r}
survvlcd4V2_2 <- survvlcd4V2_1 %>%
  # do not need to group by IndividualID because the cd4 count data variables are all already in rows with the correct IndividualId
  # create an indicator variable 'keep' to indicate if the variables from the cd4 count data set that are included in a given row of the surveillance + viral data set should be kept or not
  # Note: TestDate is the variable from the cd4 data that corresponds to when the cd4 count was measured
  # StartDate and EndDate are the dates that bound the episodes within the surveillance data set (to which viral data are merged)
  mutate(keepV2 = ifelse(((StartDate <= TestDateV2) & (TestDateV2 <= EndDate)), 1, 0),
         # if they should be kept (ie, keep == 1), then we keep the variable, 
         # and if not, then we populate that variable as NA
         TestDateV2 = ifelse(keepV2 == 1, as.Date(TestDateV2), NA), 
         CD4CountV2 = ifelse(keepV2 == 1, CD4CountV2, NA),
         CD4PctV2 = ifelse(keepV2 == 1, CD4PctV2, NA))
         
class(survvlcd4V2_2$TestDateV2) <- 'Date'

 nrow(survvlcd4V2_2)
# 2432605

# View(survvlcd4V2_2[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'VLResult', 'TestDateV2', 'CD4CountV2', 'CD4PctV2')])
```
# Step 9
Remove rows in which IndividualId, Episode, and TestDate are duplicated
```{r}
survvlcd4V2_3 <- distinct(survvlcd4V2_2, IndividualId, Episode, TestDateV2, .keep_all = TRUE) 

nrow(survvlcd4V2_3)
# 444969

# View(survvlcd4V2_3[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'VLResult', 'TestDateV2', 'CD4CountV2', 'CD4PctV2')])
# Note that some episodes are repeated because one row has the CD4 data variables while the other row is identical EXCEPT for the CD4 data variables
```
# Step 10
Within episodes (we will group by IndividualId and Episode), we will arrange by TestDate and keep only the first row of each episode group (ie, the one with the non-missing TestDate).
This addresses what we saw in the prior step in which some episodes for a given IndividualId are repeated because one row has the CD4 variables while the other row is identical EXCEPT for the CD4 data variables.
```{r}
survvlcd4V2_4 <- survvlcd4V2_3 %>%
  group_by(IndividualId, Episode) %>%
  arrange(TestDateV2) %>%
  filter(row_number() == 1) %>%
  arrange(IndividualId, Episode)

length(unique(survvlcd4V2_4$IndividualId))
# 10739

# View(survvlcd4V2_4[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'VLResult', 'TestDateV2', 'CD4CountV2', 'CD4PctV2')])
```
# Step
Create new categorical variable to categorize CD4Count; Matt suggests these 
cut-offs
```{r}
survvlcd4V2_5 <- survvlcd4V2_4 %>%
  mutate(CD4CountV2 = ifelse(CD4CountV2 >= 1800, NA, CD4CountV2),
         clinicalstageV2 = case_when(
                          (is.na(CD4CountV2)) ~ NA,
                          ((CD4CountV2 >= 500) & (CD4CountV2 < 1800)) ~ '500 or more',
                          ((CD4CountV2 >= 350) & (CD4CountV2 <= 499)) ~ '350 to 499',
                          ((CD4CountV2 >= 200) & (CD4CountV2 <= 349)) ~ '200 to 350',
                          CD4CountV2 < 200 ~ '< 200'))

survvlcd4V2_5$clinicalstageV2 <- as.factor(survvlcd4V2_5$clinicalstageV2)

####
# Check new clinicalstage variable
####
# Within person-periods in which someone is at Stage 1, which is CD4Count >= 500 cells/mm^3
check_stage500plus <- subset(survvlcd4V2_5, clinicalstageV2 == '500 or more')
# Returns Tukey's five-number summary
# min, lower hinge, median, upper hinge, max
fivenum(check_stage500plus$CD4CountV2)
# 500  566  655  793 1791
length(unique(check_stage500plus$CD4CountV2[check_stage500plus$CD4CountV2 >= 1800]))
# 0 individuals (this was 18 people before we changed anyone with counts >= 1800 to NA)

# Within person-periods in which someone is at Stage 2, which is CD4Count 350 to 499 cells/mm^3
check_stage350to499 <- subset(survvlcd4V2_5, clinicalstageV2 == '350 to 499')
# Returns Tukey's five-number summary
# min, lower hinge, median, upper hinge, max
fivenum(check_stage350to499$CD4CountV2)
# 350 384 419 458 499

# Within person-periods in which someone is at CD4Count 200 to 350 cells/mm^3
check_stage200to350 <- subset(survvlcd4V2_5, clinicalstageV2 == '200 to 350')
# Returns Tukey's five-number summary
# min, lower hinge, median, upper hinge, max
fivenum(check_stage200to350$CD4CountV2)
# 200 242 279 315 349

# Within person-periods in which someone is at CD4Count < 200 cells/mm^3
check_stageunder200 <- subset(survvlcd4V2_5, clinicalstageV2 == '< 200')
# Returns Tukey's five-number summary
# min, lower hinge, median, upper hinge, max
fivenum(check_stageunder200$CD4CountV2)
# 0  61 116 160 199

# Within person-periods in which someone is unknown
check_stageunk <- subset(survvlcd4V2_5, is.na(clinicalstageV2))
# Returns Tukey's five-number summary
# min, lower hinge, median, upper hinge, max
fivenum(check_stageunk$CD4CountV2)
# NA NA NA NA NA

#########

# View(survvlcd4V2_5[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'VisitDate', 'VLResult', 'TestDateV2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])
```
# Create window 12 months before EarliestARTInitDate and 3 months after 
# EarliestARTInitDate to create a grace period for baseline CD4 count
```{r}
# Create a flag for 12 months before EarliestARTInitDate
survvlcd4V2_6 <- survvlcd4V2_5 %>%
  group_by(IndividualId) %>%
  mutate(twelvemospriorART = (EarliestARTInitDate - dmonths(12)),
         threemospostART = (EarliestARTInitDate + dmonths(3)),
         durationdays_ARTtoCD4V2 = abs(EarliestARTInitDate - TestDateV2))

survvlcd4V2_6$durationdays_ARTtoCD4V2 <- as.numeric(survvlcd4V2_6$durationdays_ARTtoCD4V2)

# View(survvlcd4V2_6[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'twelvemospriorART', 'EarliestARTInitDate', 'threemospostART', 'TestDateV2', 'durationdays_ARTtoCD4V2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])
```

```{r}
survvlcd4V2_7 <- survvlcd4V2_6 %>%
  filter(((twelvemospriorART <= TestDateV2) & (TestDateV2 <= threemospostART))) 

# View(survvlcd4V2_7[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'twelvemospriorART', 'EarliestARTInitDate', 'threemospostART', 'TestDateV2', 'durationdays_ARTtoCD4V2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])
```
# Identify the CD4 value for each person that is closest to their baseline (their EarliestARTInit date)
```{r}
survvlcd4V2_8_test <- survvlcd4V2_7 %>%
  # Create an indicator variable to indicate the person-period with the minimum
  # duration_ARTtoCD4
  mutate(closestCD4toARTinitV2 = ifelse(durationdays_ARTtoCD4V2 == min(durationdays_ARTtoCD4V2), 1, 0),
         # if there are two CD4 values that are ties for the closestCD4toARTinitV2
         pick_row = case_when(
           sum(closestCD4toARTinitV2 == 1) > 1 & TestDateV2 > EarliestARTInitDate ~ 1, 
           sum(closestCD4toARTinitV2 == 1) == 1 ~ closestCD4toARTinitV2,
           .default = 0),
        # if it is the Episode in which the CD4 test result is closest to the
        # ART initiation (and within the filtered grace period around the 
        # EarliestARTInit date), then that's the baseline CD4 test
        base_TestDateV2 = ifelse(pick_row == 1, as.Date(TestDateV2), NA),
        base_CD4CountV2 = ifelse(pick_row == 1, CD4CountV2, NA),
        #base_clinicalstageV2 = ifelse(closestCD4toARTinitV2 == 1, as.factor(clinicalstageV2), NA),
        base_CD4PctV2 = ifelse(pick_row == 1, CD4PctV2, NA),
        base_clinicalstageV2 = case_when(
                          (!is.na(base_TestDateV2)) & ((base_CD4CountV2 >= 500) & (base_CD4CountV2 < 1800)) ~ '500 or more',
                          (!is.na(base_TestDateV2)) & ((base_CD4CountV2 >= 350) & (base_CD4CountV2 <= 499)) ~ '350 to 499',
                          (!is.na(base_TestDateV2)) & ((base_CD4CountV2 >= 200) & (base_CD4CountV2 <= 349)) ~ '200 to 350',
                          (!is.na(base_TestDateV2)) & (base_CD4CountV2 < 200) ~ '< 200'))

class(survvlcd4V2_8_test$base_TestDateV2) <- 'Date'
survvlcd4V2_8_test$clinicalstageV2 <- as.factor(survvlcd4V2_8_test$clinicalstageV2)
nrow(survvlcd4V2_8_test)
#7665
length(unique(survvlcd4V2_8_test$IndividualId))
# 4342

#View(survvlcd4V2_8_test[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'TestDateV2', 'durationdays_ARTtoCD4V2', 'closestCD4toARTinitV2', 'pick_row', 'base_TestDateV2', 'base_CD4CountV2', 'base_CD4PctV2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])

#View(survvlcd4V2_8_test[survvlcd4V2_8_test$IndividualId == '66271', c('IndividualId', 'Episode', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'pick_row', 'base_TestDateV2', 'base_CD4CountV2')])

#View(survvlcd4V2_8_test[survvlcd4V2_8_test$IndividualId == '139718', c('IndividualId', 'Episode', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'pick_row', 'base_TestDateV2', 'base_CD4CountV2')])
```
Filter to all rows with non-missing baseline CD4 values
```{r}
length(unique(survvlcd4V2_8_test$IndividualId))
# 4342

survvlcd4V2_9 <- survvlcd4V2_8_test %>%
  filter(!is.na(base_CD4CountV2))

# View(survvlcd4V2_9[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'twelvemospriorART', 'EarliestARTInitDate', 'threemospostART', 'TestDateV2', 'durationdays_ARTtoCD4V2', 'closestCD4toARTinitV2', 'base_TestDateV2', 'base_CD4CountV2', 'base_CD4PctV2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])
```
Restrict to the baseline CD4 variables
```{r}
survvlcd4V2_10 <- survvlcd4V2_9 %>%
  select(IndividualId, durationdays_ARTtoCD4V2, base_TestDateV2, base_CD4CountV2, base_clinicalstageV2, base_CD4PctV2)

# View(survvlcd4V2_10)
# Note that some people have a CD4Count for a person-time in which they have a NA CD4Pct
 
#View(survvlcd4V2_10[survvlcd4V2_10$IndividualId == '66271', ])

#View(survvlcd4V2_10[survvlcd4V2_10$IndividualId == '139718', ])

```
Now merge it back
```{r}
survvlcd4V2_11 <- left_join(x = survvlcd4V2_5, y = survvlcd4V2_10, by = 'IndividualId')

# View(survvlcd4V2_11[ , c('IndividualId', 'Episode', 'StartDate', 'EndDate', 'EarliestARTInitDate', 'TestDateV2', 'durationdays_ARTtoCD4V2', 'base_TestDateV2', 'base_CD4CountV2', 'base_clinicalstageV2', 'base_CD4PctV2', 'CD4CountV2', 'clinicalstageV2', 'CD4PctV2')])
```
########################################
# Remember we called the data set df6 before we went back and merged in the cd4 count data
# Here, we will substitute survvlcd4_11 for df6, and then continue on with df7, which will now include the fully merged data set with surveillance+viral+cd5
```{r}
######
# After we dropped the persons who had the event before EarliestARTInitDate
######

length(unique(survvlcd4V2_11$IndividualId[survvlcd4V2_11$suppression_pre_ART == 1]))
# 0 confirms that we dropped everyone who had the event before EarliestARTInitDate

length(unique(survvlcd4V2_11$IndividualId[survvlcd4V2_11$suppression_pre_ART == 0]))
# 10739

# Remember we dropped 242 individuals who experienced viral control before EarliestARTInitDate
# Keep these numbers for the STROBE
# 10981-242=10739
```
# Step 15
Create origin for each person
```{r}
df7 <- survvlcd4V2_11 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # Create variable to indicate the origin for the survival analysis
  # Remember: EarliestARTInitDate is not always the same as an episode StartDate 
  mutate(origin = ifelse(((EarliestARTInitDate >= StartDate) & (EarliestARTInitDate <= EndDate)), 1, 0)) %>%
  mutate(origins_exists_test = ifelse(sum(origin == 1) > 1, 1, 0)) %>%
  mutate(origin_exists_test = ifelse(sum(origin == 1) == 1, 1, 0))

# Check that no one has a missing value for origin or origin_exists
length(unique(df7$IndividualId[is.na(df7$origin)]))
# 0
length(unique(df7$IndividualId[is.na(df7$origins_exists_test)]))
# 0
length(unique(df7$IndividualId[is.na(df7$origin_exists_test)]))
# 0
# No one has NA for any of these variables

# How many people had an origin within their episodes
length(unique(df7$IndividualId[df7$origin == 1]))
# 5446
sum(df7$origin == 1)
# 5446

# How many people had more than one origin
length(unique(df7$IndividualId[df7$origins_exists_test == 1]))
# 0

# Confirm Episode 75 is no longer repeated for IndiviudalId 66271
#View(df7[df7$IndividualId == '66271' & df7$origin == 1, c('IndividualId', 'Episode', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'base_TestDateV2', 'base_CD4CountV2', 'origin', 'origin_exists_test')])

# How many people had only one origin
length(unique(df7$IndividualId[df7$origin_exists_test == 0]))
# 5293
length(df7$IndividualId[df7$origin == 1])
# 5446
# It makes sense that even though we subsetted the data to include people whose 
# EarliestARTInitDate fell within 2011-2018, there is no guarantee that they 
# have person-time that extends to that EarliestARTInitDate
# Remember that EarliestARTInitDate comes from Tier.net. and surveillance and 
# viral load data come from AHRI

# View(df7[, c('IndividualId', 'Episode', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'origin', 'origin_exists_test')])
```
# Step 16
Subset to persons who have an origin (some amount of FUP time in which they started ART and then could have gotten the outcome)
```{r}
df8 <- df7 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  filter(origin_exists_test == 1) 

length(unique(df8$IndividualId[df8$origin_exists_test == 1]))
# 5446
length(unique(df8$IndividualId))
# 5446
length(unique(df8$IndividualId[df8$origin_exists_test == 0]))
# 0

# nrow(df8) #means nothing because we still have rows from before EarliestARTInitDate
# 350484

# View(df8[, c('IndividualId', 'Episode', 'Resident', 'StartDate', 'EarliestARTInitDate', 'EndDate', 'origin', 'origin_exists_test')])
```
# Step 16b | STROBE
How do we go from 10739 to 5446?
```{r}
length(unique(df7$IndividualId[df7$origin_exists_test == 1]))
# 5446 persons had an origin included in their person-time

length(unique(df7$IndividualId[df7$origin_exists_test == 0]))
# 5293 persons did not have person-time during the follow-up period

# 10739-5293
# 5446 persons had an origin within their person-time during the FUP period
```
# Step 17
Create indicator of person-time 1 year prior to EarliestARTInitDate
```{r}
# Calculate one year prior to EarliestARTInitDate
df9 <- df8 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # Create a variable to identify the date 12 months prior to EarliestARTInitDate
  mutate(oneyr_preART_date = EarliestARTInitDate %m-% days(365),
        # Create an indicator to flag the episode that contains the start of that 12-month period
        oneyr_preART_start = ifelse(((oneyr_preART_date >= StartDate) & (oneyr_preART_date <= EndDate)), 1, 0),
        # Create a time invariant variable to indicate if someone has an exposure start 
        oneyr_preART_start_exist = ifelse(length(which(oneyr_preART_start == 1)) == 1, 1, 0)) 

length(unique(df9$IndividualId[df9$oneyr_preART_start_exist == 1]))
# 5052 persons have person-time a year prior to their EarliestARTInitDate

length(unique(df9$IndividualId[df9$oneyr_preART_start_exist == 0]))
# 394 persons have no exposure start indicator 
```
# Step 18
Restrict to persons who had person-time 1 year prior to their EarliestARTInitDate
```{r}
df10 <- df9 %>% 
  # Now restrict analytic sample to persons who had person-time a year before they initiated ART
  filter(oneyr_preART_start_exist == 1) 

length(unique(df10$IndividualId))
# 5052

# Check if everyone within the subset has either a value 1 or 0 for a migration history start (important because if you omit '=' from '>=' above for filtering by EndDate, it changes this)
length(df10$IndividualId[df10$oneyr_preART_start == 1])
# 5052
length(unique(df10$IndividualId[df10$oneyr_preART_start_exist == 0]))
# 0 persons have no migration history start indicator because we have filtered to those who have one

print(df10$IndividualId[df10$oneyr_preART_start_exist == 1 & sum(df10$oneyr_preART_start) < 1])
# no results - also what we expect and need here

# View(df10[, c('IndividualId', 'Episode', 'EarliestARTInitDate', 'StartDate', 'oneyr_preART_date', 'oneyr_preART_start', 'EndDate', 'oneyr_preART_start_exist')])
```
# Step 18b | STROBE
How do we go from 5446 to 5052?
```{r}
# Using data from before we restricted to persons who had person-time that contains the date that is 12 months before EarliestARTInitDate
length(unique(df9$IndividualId[df9$oneyr_preART_start_exist == 1]))
# 5052 persons have person-time a year prior to their EarliestARTInitDate

length(unique(df9$IndividualId[df9$oneyr_preART_start_exist == 0]))
# 394 persons do not have person-time that contains the date that is 12 months before EarliestARTInitDate 

# 5446-394
# 5052 persons have person-time a year prior to their EarliestARTInitDate
```
# Step 19
Create indicator to flag person-time 1 day prior to EarliestARTInitDate
```{r}
df11 <- df10 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # Create end of one-year pre-ART person-time as one day before EarliestARTInitDate
  mutate(oneday_preART_date = (ymd(EarliestARTInitDate) - days(1)),
        # Create indicator to flag the episode with the end of the 1-year person-time before EarliestARTInitDate
        oneday_preART_end = ifelse(((oneday_preART_date >= StartDate) & (oneday_preART_date <= EndDate)), 1, 0),
        oneday_preART_end_exist = ifelse((length(which(oneday_preART_end == 1)) == 1), 1, 0),
        # Calculate the elapsed time between each person's start and end date of one year prior to EarliestARTInit
        oneyr_preART_duration_days = ((oneyr_preART_date %--% oneday_preART_date) %/% days(1)))
  
# Confirm we have isolated one year of person-time pre-ART
summary(df11$oneyr_preART_duration_days)
   # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   #  364     364     364     364     364     364 

length(unique(df11$IndividualId))
# 5052
length(unique(df11$IndividualId[df11$oneyr_preART_start == 1]))
# 5052
length(unique(df11$IndividualId[df11$oneyr_preART_start_exist == 1]))
# 5052
length(unique(df11$IndividualId[df11$oneday_preART_end == 1]))
# 5052
length(unique(df11$IndividualId[df11$oneday_preART_end_exist == 1]))
# 5052

# View(df11[ , c('IndividualId', 'Episode', 'StartDate', 'oneyr_preART_date', 'oneday_preART_end_date', 'EndDate', 'oneyr_preART_start', 'oneyr_preART_start_exist', 'oneday_preART_end', 'oneday_preART_end_exist')])
```
# Step 20
Create migration history covariate by calculating the number of migration events during the year prior to EarliestARTInitDate
```{r}
df12 <- df11 %>% 
  group_by(IndividualId) %>% 
  arrange(IndividualId, Episode) %>%
  # Filter to episodes that fall within one year of person-time before EarliestARTInit
  # Use >= and <= as the operators below, respectively, because some persons have their oneyr_preART_date such that the episode is only 1 day in duration (in which case, StartDate, oneyr_preART_date, and EndDate are all the same). 
  filter((EndDate >= oneyr_preART_date) & (EndDate <= oneday_preART_date) & (oneyr_preART_start_exist == 1) & (oneday_preART_end_exist == 1)) %>%
  # Create summary of median household SEIdx during the year before ART start (one-year history of SEIdx before ART start)
  # Include na.rm = TRUE so that we drop the NAs from the observations during that year before calculating the median (we cannot calculate the median of a set of numbers that includes NA values)
  mutate(SEIdx_hx = median(SEIdx, na.rm = TRUE)) %>%
  # Create flag for instances in which Resident==1 is followed by Resident==0, indicating an out-migration event
  mutate(outmig_hx_flag = if_else(((Lag(Resident) == 1) & (Resident == 0)), 1, 0, 0),
  # Exposure history definition 1: frequency of out-migration events per person
        outmig_hx_freq = length(which(outmig_hx_flag == 1)),
  # Exposure history definition 2: pattern of out-migration per person
        outmig_hx_freq_cat = case_when(
            # always out-migrant
              # If someone has 0 episodes in which they are home (Resident == 1); they are never home, and always an out-migrant
            length(which(Resident == 1)) == 0 ~ 2, 
            # sometimes out-migrant
              # If someone has 1 or more out-migration event flags, they are out-migrating >= times that year
            (length(which(outmig_hx_flag == 1)) >= 1) ~ 1,
            # never out-migrant (always permanent resident, non-migrant)
              # If someone has Resident == 1 for all episodes within this time frame OR zero out-migration events (eg, 0, 0, 1)
            ((length(which(Resident == 1)) == length(IndividualId)) | (length(which(outmig_hx_flag == 1)) == 0)) ~ 0
        )
  )
            
# View(df12[, c('IndividualId', 'Episode', 'Resident', 'oneyr_preART_start', 'oneday_preART_end', 'outmig_hx_flag', 'outmig_hx_freq', 'outmig_hx_freq_cat', 'SEIdx', 'SEIdx_hx', 'StartDate', 'EndDate', 'VisitDate', 'oneyr_preART_start_exist', 'oneday_preART_end_exist')])
 
# Note: SEIdx_hx variable, designed to take the median of all SEIdx in the year before
 # ART initiation, so that we can carry it forward as baseline SEIdx in cases in which 
 # SEI is missing at baseline - this variable doesn't solve all problems because
 # there may be some individuals (like IndividualId==980) who are missing SEIdx
 # for the entirety of the year before ART start
```
# Step 21
Within this filtered data frame with the migration variables, retain only the IndividualId, migration frequency history variable, and migration pattern history variable, plus the new SEIdx history variable (time-invariant baseline covariates)
```{r}
df13 <- df12 %>%
  select(IndividualId, outmig_hx_freq, outmig_hx_freq_cat, SEIdx_hx)

# View(df13)
```
# Step 22
Within the filtered data frame with the migration variables, retain only unique rows 
-Note that the migration frequency history variable and migration pattern history variables are time-invariant by definition, so we only need one value per IndividualId
-Want one row per IndividualId in order to do a one-to-many merge by IndividualId
```{r}
# We want to keep all variables in this data set
df14 <- distinct(df13, .keep_all = TRUE)
# View(df14)
```
# Step 22
Left-join the surveillance data set (x=pre-filter data frame) to this filtered data frame with the migration history variable (y=migration history variable data frame) by IndividualId
```{r}
df15 <- left_join(x = df11, y = df14, by = 'IndividualId')
# View(df15)
# View(df15[, c('IndividualId', 'Episode', 'Resident', 'oneyr_preART_start', 'oneday_preART_end', 'outmig_hx_freq', 'outmig_hx_freq_cat', 'SEIdx', 'SEIdx_hx', 'StartDate', 'EndDate', 'VisitDate', 'oneyr_preART_start_exist', 'oneday_preART_end_exist')])
```
# Step 23
Select only columns we need moving forward
```{r}
df16 <- df15 %>%
  select(IndividualId, Episode, LocationId, StartDate, EndDate, Resident, EarliestARTInitDate, DoD, Died, Age, SEIdx, SEIdx_hx, Sex, edu, marital, ChildNumber, oneyr_preART_start, oneday_preART_end, outmig_hx_freq, outmig_hx_freq_cat, VisitDate, BMFStatusObservation, HIVSampleId, VLDetectableLimit, VLResult, VLResultCode, TestDateV2, CD4CountV2, clinicalstageV2, CD4PctV2, base_TestDateV2, base_CD4CountV2, base_clinicalstageV2, base_CD4PctV2, durationdays_ARTtoCD4V2, suppression_pre_ART, origin, origin_exists_test, elapsed.time, Year, Cause1, Lik1, Cause2, Lik2, Cause3, Lik3, ImmediateCause, UnderlyingCause, AssociatedCause1, AssociatedCause2, LostToFollowUp, Refusal, Education, MaritalStatus)
```
# Step 24
Create an indicator to flag when someone's VisitDate occurred before someone's EarliestARTInitDate
```{r}
df17 <- df16 %>%
  group_by(IndividualId, VisitDate) %>%
  arrange(IndividualId, Episode) %>%
  mutate(drop_visit = ifelse(VisitDate <= EarliestARTInitDate, 1, 0))
# View(df17[, c('IndividualId', 'Episode', 'Resident', 'StartDate', 'EndDate', 'VisitDate', 'EarliestARTInitDate', 'drop_visit')])
```
# Step 25
Set any VisitDates that came before or on EarliestARTInitDate == NA so that they cannot be evaluated for events 
-We already removed people who had the event pre-EarliestARTInitDate
```{r}
df18 <- df17 %>%
  mutate(VisitDate = ifelse(drop_visit == 1, NA, VisitDate), 
         BMFStatusObservation = ifelse(drop_visit == 1, NA, BMFStatusObservation), 
         HIVSampleId = ifelse(drop_visit == 1, NA, HIVSampleId), 
         VLDetectableLimit = ifelse(drop_visit == 1, NA, VLDetectableLimit), 
         VLResult = ifelse(drop_visit == 1, NA, VLResult),
         VLResultCode = ifelse(drop_visit == 1, NA, VLResultCode)
         )

class(df18$VisitDate) <- 'Date'
# View(df18[, c('IndividualId', 'Episode', 'Resident', 'StartDate', 'EndDate', 'VisitDate', 'EarliestARTInitDate', 'drop_visit')])

# Check that every row which was flagged as having a VisitDate before EarliestARTInitDate now has VisitDate == NA
# testing <- subset(df18, drop_visit == 1)
# View(testing[, c('IndividualId', 'Episode', 'Resident', 'StartDate', 'EndDate', 'VisitDate', 'EarliestARTInitDate', 'drop_visit')])
# length(which(!is.na(testing$VisitDate)))
# 0 persons have a non-missing VisitDate if the VisitDate was flagged as occurring before EarliestARTInitDate

# Check another way
# test <- df18 %>%
#   group_by(IndividualId) %>%
#   arrange(IndividualId, Episode) %>%
#   mutate(error = ifelse((!is.na(VisitDate))&(!is.na(EarliestARTInitDate))&(VisitDate <= EarliestARTInitDate), 1, 0))
# 
# length(unique(test$IndividualId[test$error == 1]))
# 0
```
# Step 26
Create binary viral suppression variable
```{r}
df19 <- df18 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # str(df18$VLResultCode) confirms that there are 7 levels, as indicated in the data dictionary
  # Leave NAs as NA because we do not want to collapse viral failure values with NAs (ie, do not replace NAs with 0s) 
  # Create a binary variable to distinguish a viral result that shows suppression versus one that does not
  mutate(suppression = if_else(VLResultCode == 2, 1, 0, NA_real_)) 

length(unique(df19$IndividualId))
# 5052
# View(df19[, c('IndividualId', 'Episode', 'origin', 'VisitDate', 'EarliestARTInitDate', 'VLResultCode', 'suppression', 'StartDate', 'EndDate')])
```
# Step 27
Create binary initial viral suppression variable
```{r}
df20 <- df19 %>%
  group_by(IndividualId) %>%
  arrange(IndividualId, Episode) %>%
  # Create a flag for each time suppression == 1
  mutate(suppression_flag = if_else(suppression == 1, 1, NA_real_, NA_real_),
  # Rank the flags for each time suppression == 1
  # We 1) rank only the suppression == 1 values, 2) keep NAs as NA (do not rank NAs) 3) give a rank of 0 to all suppression == 0 values
        suppression_rank = ifelse(suppression == 1, rank(suppression_flag, na.last = 'keep', ties.method = 'first'), NA_real_),
  # Create binary event variable to indicate initial viral suppression
        event_initsup = case_when(
                # Initial viral suppression is the event
                suppression_rank == 1 ~ 1,
                # Viral failure, or viral suppression that occurs after the initial suppression is not an event
                ((suppression == 0) | (suppression_rank > 1)) ~ 0
  )
  )
  
 length(unique(df20$IndividualId))
# 5052
dim(df20)
#  343260     59
# View(df20[, c('IndividualId', 'Episode', 'origin', 'VisitDate', 'EarliestARTInitDate', 'VLResultCode', 'suppression', 'suppression_flag', 'suppression_rank', 'event_initsup')])
```
# Step 28
Create assessment_date, defined as EarliestARTInitDate when an episode contains the origin, which is EarliestARTInitDate, and as the VisitDate when we have a VisitDate
```{r}
# df21 <- df20 %>%
#   group_by(IndividualId) %>%
#   arrange(IndividualId, Episode) %>%
#   # Create an assessment date that is the date of EarliestARTInitDate and then any VisitDates (we are counting all as 'assessments')
#   mutate(assessment_date = case_when(
#     # If VisitDate is missing and an episode contains the origin, which is EarliestARTInitDate, assessment_date == EarliestARTInitDate
#     ((is.na(VisitDate)) & (origin == 1)) ~ EarliestARTInitDate,
#     # If VisitDate is not missing, assessment_date == VisitDate
#     !is.na(VisitDate) ~ VisitDate))
# 
# class(df21$assessment_date) <- 'Date'

# length(unique(df21$IndividualId))
# 5052

# View(df21[, c('IndividualId', 'Episode', 'origin', 'VisitDate', 'EarliestARTInitDate', 'event_initsup', 'assessment_date')])
```
# Step 29
Create assessment_rank to rank the assessment_date 
```{r}
# df22 <- df21 %>%
#   group_by(IndividualId) %>%
#   arrange(IndividualId, Episode) %>%
#   # Create a flag to indicate if a row contains the assessment date (if it is part of the person-assessment backbone)
#   # If the assessment_date falls within a given episode, we flag that episode as being part of the person-assessment backbone
#   mutate(backbone_flag = ifelse(((StartDate <= assessment_date) & (EndDate >= assessment_date)), 1, NA),
#     # Create assessment_rank to number the assessments within a given person
#     # If a person-assessment has been flagged as falling within an episode (ie, backbone_flag == 1), we count the number of flags
#     assessment_rank = ifelse(((backbone_flag == 1) & (!is.na(backbone_flag))), rank(backbone_flag, na.last = TRUE, ties.method = 'first'), NA))
  
# length(unique(df22$IndividualId))
# 5052

# View(df22[, c('IndividualId', 'Episode', 'origin', 'VisitDate', 'EarliestARTInitDate', 'event_initsup', 'assessment_date', 'backbone_flag', 'assessment_rank')])
```
Explore data on earliest known ART initiation, and how many participants started before/after UTT
```{r}
# check what proportion of participants started ART before UTT policy was put into place 
# officially rolled out Sept 01, 2016 per publication
df21 <- df20 %>%
  # Create binary variable to indicate if someone initiated before UTT
  mutate(EarliestARTInitPreUTT = as.factor(ifelse(as.Date(EarliestARTInitDate, '%Y/%m/%d') < as.Date('2016-09-01'), 1, 0)),
         # Create variable to indicate year of someone's ART initiation
         EarliestARTInitYear = year(EarliestARTInitDate))

# Check distribution of year of EarliestARTInitDate
df21_baseline <- df21 %>%
  group_by(IndividualId) %>%
  slice_head(n = 1)

sum(is.na(df21_baseline$EarliestARTInitYear))
# 0

####
# Explore data to get counts for bar chart
####
table(df21_baseline$EarliestARTInitYear, exclude = NULL)
# 2011 2012 2013 2014 2015 2016 2017 2018 
# 953  934  805  849 1045  374   29   63 

sum((df21_baseline$EarliestARTInitYear == 2016) & (df21_baseline$EarliestARTInitPreUTT == 1))
# 346 individuals started ART in 2016 before Sept 01, 2016

sum((df21_baseline$EarliestARTInitYear == 2016) & (df21_baseline$EarliestARTInitPreUTT == 0))
# 28 individuals start ART in 2016 on or after Sept 01, 2016

# raw data from above
year <- c(2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018)
count <- c(953, 934, 805, 849, 1045, 374, 29, 63)

# We can see what the histogram should look like, when we clean it up below
hist(df21_baseline$EarliestARTInitYear)

# Create data frame for stacked bar chart of distribution of ART initiation
year <- c(2011, 2011, 2012, 2012, 2013, 2013, 2014, 2014, 2015, 2015, 2016, 2016, 2017, 2017, 2018, 2018)
count <- c(953, 0, 934, 0, 805, 0, 849, 0, 1045, 0, 346, 28, 0, 29, 0, 63)
preUTT <- rep(c("Pre-Universal Test & Treat Policy", "Post-Universal Test & Treat Policy"), 8)
data_graph <- data.frame(year, count, preUTT)
data_graph$preUTT <- factor(data_graph$preUTT, levels = c('Pre-Universal Test & Treat Policy', 'Post-Universal Test & Treat Policy'))

# Create stacked bar chart of distribution of ART initiation
Fig_ARTInit <- ggplot(data = data_graph, 
                      aes(x = year,
                          y = count,
                          fill = preUTT)) +
                geom_bar(position = 'stack', stat = 'identity', width = .8) + 
                labs(x = 'Year of earliest known antiretroviral therapy initiation', 
                     y = str_wrap('Number of participants', width = 10)) + 
                theme_light() +
                # Set limits and intervals for x- and y-axes
                scale_x_continuous(limits = c(2011, 2018), 
                                   breaks = seq(from=2011, to=2018, by=1),
                                    oob = rescale_none) +
                scale_fill_manual(values = c('#D72000FF', '#FFAD0AFF')) +
                # scale_fill_paletteer_d('LaCroixColoR::Apricot') +
                theme(legend.position = 'bottom',
                      legend.title = element_blank(),
                      text = element_text(family = "Times New Roman"),
                      axis.title.y = element_text(angle = 0, vjust = 0.5, 
                      # Add space between y-axis label and y-axis                       
                      margin = margin(t = 0, r = 10, b = 0, l = 0)),
                      axis.title.x = element_text(
                        margin = margin(t = 10, r = 0, b = 0, l = 0))) 
Fig_ARTInit
```
# Step 30
Subset to baseline (ie, time of ART initiation) observations for Table 1
```{r}
# Subset to the origin row for each person (ie, at EarliestARTInitDate)
# It's okay if this row contains person-time before EarliestARTInitDate because the covariates have not changed within that episode by definition
baseline <- subset(df21, origin == 1)

# length(unique(baseline$IndividualId))
# 5052

# length(baseline$IndividualId[baseline$origin == 1])
# 5052

# View(baseline[, c('IndividualId', 'Episode', 'origin', 'outmig_hx_freq', 'outmig_hx_freq_cat', 'Age', 'Sex', 'edu', 'marital', 'SEIdx', 'event_initsup')])
```
Code to use for exporting list of individuals in order to pull their CD4 count data
```{r}
#sample <- subset(baseline, select = c('IndividualId', 'EarliestARTInitDate'))
#sample <- as.data.frame(sample)
#write_xlsx(sample, 'Table_migration_mortality_study_sample.xlsx')
```
# Step 31
Create Table 1 of baseline (ie, time of ART initiation) sociodemographic characteristics
```{r}
baseline$ChildNumber <- as.numeric(baseline$ChildNumber)
baseline$outmig_hx_freq_cat <- as.factor(baseline$outmig_hx_freq_cat)
baseline$Sex <- droplevels(baseline$Sex)

table1 <- CreateTableOne(vars = c('Age', 'edu', 'marital', 'SEIdx', 'ChildNumber', 'outmig_hx_freq_cat', 'SEIdx_hx'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)

table1 <- print(table1, showAllLevels = TRUE, includeNA = TRUE)
table1 <- as.data.frame(table1)
write_xlsx(table1, 'Table 1_SociodemographicCharacteristics.xlsx')
```
Check distributions of variables that have baselinemissingness
```{r}
table_baseline_edu <- CreateTableOne(vars = c('Education', 'edu'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)
table_baseline_edu
```

```{r}
table_baseline_sei <- CreateTableOne(vars = c('SEIdx', 'SEIdx_hx'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)

print(table_baseline_sei, nonnormal = c('SEIdx', 'SEIdx_hx'))

sum(is.na(baseline$SEIdx))
# 289/5052 = 5.7%

fivenum(baseline$SEIdx)
```

```{r}
table_baseline_cohab <- CreateTableOne(vars = c('MaritalStatus', 'marital'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)

table_baseline_cohab
```

```{r}
table_baseline_cd4 <- CreateTableOne(vars = c('base_CD4CountV2', 'base_clinicalstageV2'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)

print(table_baseline_cd4, nonnormal = c('base_CD4CountV2'))

fivenum(baseline$base_CD4CountV2)
```
Create Table 2 of baseline (ie, time of ART initiation) clinical characteristics
```{r}
table2 <- CreateTableOne(vars = c('EarliestARTInitPreUTT', 'durationdays_ARTtoCD4V2', 'base_clinicalstageV2', 'base_CD4CountV2', 'base_CD4PctV2'), strata = 'Sex', data = baseline, includeNA = TRUE, addOverall = TRUE)

table2 <- print(table2, nonnormal = c('durationdays_ARTtoCD4V2', 'base_CD4CountV2', 'base_CD4PctV2'), showAllLevels = TRUE, includeNA = TRUE)
table2 <- as.data.frame(table2)
write_xlsx(table2, 'Table 2_ClinicalCharacteristics.xlsx')
```
Missingness for clinical characteristics
```{r}
# CD4Count
length(which(is.na(baseline$base_CD4CountV2)))
# 711
length(which(is.na(baseline$base_CD4CountV2) & (baseline$Sex == 'Male')))
# 152
length(which(is.na(baseline$base_CD4CountV2) & (baseline$Sex == 'Female')))
# 559
length(which(!is.na(baseline$base_CD4CountV2)))
# 4341

# TestDate
length(which(is.na(baseline$base_TestDateV2)))
# 711
length(which(is.na(baseline$base_TestDateV2) & (baseline$Sex == 'Male')))
# 152
length(which(is.na(baseline$base_TestDateV2) & (baseline$Sex == 'Female')))
# 559
length(which(!is.na(baseline$base_TestDateV2)))
# 4341

# duration
length(which(is.na(baseline$durationdays_ARTtoCD4V2)))
# 711
length(which(is.na(baseline$durationdays_ARTtoCD4V2) & (baseline$Sex == 'Male')))
# 152
length(which(is.na(baseline$durationdays_ARTtoCD4V2) & (baseline$Sex == 'Female')))
# 559
length(which(!is.na(baseline$durationdays_ARTtoCD4V2)))
# 4341

# CD4Pct
length(which(is.na(baseline$base_CD4PctV2)))
# 3836
length(which(is.na(baseline$base_CD4PctV2) & (baseline$Sex == 'Male')))
# 1060
length(which(is.na(baseline$base_CD4PctV2) & (baseline$Sex == 'Female')))
# 2776
length(which(!is.na(baseline$base_CD4PctV2)))
# 1216
```

Alternative Table 1, stratified by out-migration history, of baseline (ie, time of ART initiation) sociodemographic characteristics
```{r}
baseline$ChildNumber <- as.numeric(baseline$ChildNumber)

t <- CreateTableOne(vars = c('Age', 'Sex', 'edu', 'marital', 'SEIdx', 'ChildNumber'), strata = 'outmig_hx_freq_cat', data = baseline, includeNA = TRUE, addOverall = TRUE)

t <- print(t, showAllLevels = TRUE, includeNA = TRUE)
t <- as.data.frame(t)
write_xlsx(t, 'Table 1_SociodemographicsByMigrationHx.xlsx')
```
Summarise baseline CD4 values
```{r}
table(baseline$base_CD4CountV2)
```
# Step 32
Save the data frame with clean sociodemographic data for those retained within the analytic sample
```{r}
Step02AnalyticSample <- df21
save(Step02AnalyticSample, file = '02_AnalyticSample.Rdata')
```